---
title: "New Tools for Data Analysis"
author: "Jonathan Gilligan"
date: "10/20/2018"
fontfamily: "newtxtext,newtxmath"
fontsize: "12pt"
output:
  pdf_document: default
  html_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(broom)
library(magrittr)
library(kayadata)

theme_set(theme_bw())
```
# Some New Tools for Working with Data

There are two new tools we will use in working with data for this lab:

## Getting fancy with the pipe operator (`%>%`)

With the tidyverse, we often use the "pipe" operator `%>%` to send the
output from one function into the input of another function. In most cases,
this works very simply and it uses the output from the function on the
left as the first argument of the function on the right. 

For instance, the `filter` function selects certain rows in a data frame or
tibble. I will work here with examples from the built-in `storms` data
frame that lists a subset of hurricanes from the National Oceanic and 
Atmospheric Administration's Atlantic hurricane database.
The data includes the positions and attributes of 198 tropical storms, measured
every six hours during the lifetime of the storm.

Let's select Hurricane Katrina from 2005 (I specify the year because there
were two earlier Katrinas in 1981 and 1999; storm names are re-used 
periodically unless they are retired because a storm with that name becomes 
historically important).
```r
data(storms)
katrina <- filter(storms, name == "Katrina", year == 2005)
```
and we can also do the same thing using the pipe operator like this:
```r
storms %>% filter(name == "Katrina", year == 2005)
```
What the pipe operator `%>%` does behind the scenes is take that output of
whatever is on the left (the data frame `storms`) and stick it in front
of the arguemnts for the function on the right so the two examples given
above are equivalent.

We have been usuing the pipe operator throughout the semester to combine 
functions. 
Suppose we want to `mutate` the `storms` data frame to add a new 
column `ts_area`
that is equal to `pi * (ts_diameter / 2)^2`, where `ts_diameter` is a column
in `storms` that corresponds to the diameter (in nautical miles) of the area 
in which winds are tropical storm strength or greater (36 knots), 
and then we want to select only the data where the tropical storm area is 
greater than 10,000 square nautical miles, and finally sort the data so the
rows of the data frame go from largest to smallest tropical storm area.
We can do that as follows:
```r
# First way
df1 = arrange(filter(mutate(storms, ts_area = pi * (ts_diameter / 2.0)^2), 
     ts_area > 10000), desc(ts_area))

# Second way, using a temporary variable
df2 = mutate(storms, ts_area = pi * (ts_diameter / 2.0)^2)
df2 = filter(df2, ts_area > 10000)
df2 = arrange(df2, desc(ts_area))

# Third way, using the pipe operator
df3 = storms %>% mutate(storms, ts_area = pi * (ts_diameter / 2.0)^2) %>%
     filter(ts_area > 10000) %>% arrange(desc(ts_area))
```
We use pipes because often they are simpler to understand. If we use the first
way in the example above, if we are combining many functions, nesting functions
inside other functions becomes very difficult to read and understand.
The second way, of using temporary variables is fine, but it can get confusing 
to have lots of variable names that we use only one time to pass data from
one function to another. The third way, using the pipe operator makes it easier
to understand how the data is moving through the processing pipeline.

But the way pipe operators work can sometimes get in the way of things we want
to do.

First, suppose that I want to send the results of a function into the second
or third argument of a subsequent function instead of the first argument?
We use the `lm` function to calculate the slope and intercept of a trend line
that best fits some data. The lm function expects the formula for the line
to fit to be the first argument and the data to be the second.

Suppose I want to examine hurricanes from  the year 2005 and look for a 
relationship between the air pressure in the eye and the wind speed,
but only during the time when they are classified as hurricanes.

I can do this as follows:
```r
hurricanes_2005 = storms %>% filter(status == "hurricane", year == 2005)
lin_fit = lm(wind ~ pressure, data = hurricanes_2005)
```
The `lm` function looks at the data given by the `data` argument and
then finds the best line that characterizes the linear relationship
`wind ~ pressure`, which means that the wind is a linear function of the
pressure (this would be the slope of a graph where `wind` is on the y-axis
and `pressure` is on the x-axis).

Suppose that I would like to pipe data from the `filter` function to the
`lm` function. I have a problem because the pipe operator would put the
output from `filter` as the first argument of the `lm` function, but in
`lm` the first argument is the formula and `data` is the second.
I can do this as follows:
```r
lin_fit2 = storms %>% filter(status == "hurricane", year == 2005) %>%
           lm(wind ~ pressure, data = .)
```
The pipe operator creates a new variable `.` (the period), which gets the
value of whatever is on the left of the operator. If I don't use the `.`
in the function on the right, the pipe operator invisibly puts `.` in
the first argument so `foo(x) %>% bar(y, z)` is re-written behind the 
scenes to become
```r
. = foo(x)
bar(., y, z)
```
However, if I want to use the `.` somewhere else, I can do that explicitly
and when the pipe operator looks at the function on its right, it knows
not to stick the `.` in the first argument if I use `.` in the function
call. This, I could write `foo(y) %>% bar(x, ., z)` if I want to substitute
the output of `foo(y)` for the second argument of `bar`.

## Extracting columns from a data frame in a pipeline

If I just want to look at the names of the hurricanes from the `storms` data
frame, I can use the `$` operator:
```r
names = storms$name
```
But suppose that I want to filter the storms and then extract the names.
What are the names of category 5 hurricanes?
```r
cat_5 = storms %>% filter(category == 5)
names = cat_5$name
```
Suppose I want to do this without creating a variable `cat_5`?
I could try 
```r
# This will give an error
names = storms %>% filter(category == 5)$name
```
but it turns out that this doesn't work because of the way the pipe operators
work.
To do this, I need to load the `magrittr` library and use the `%$%` operator
instead of `$`:
```r
# This will work
names = storms %>% filter(category == 5) %$% name
```
The `%$%` operator works like the pipe operator, but instead of calling a function,
it extracts the specified column (in this case, `name`).

This will be useful for what comes next.

## Tidying up results of statistical modeling

`lm` is one example of a lot of powerful functions in R that let you analyze
data by fitting models. A statistical model is an idealized functional relationship
between different attributes of the data (attributes are represented by columns
in a data frame). 
For instance, a linear model idealizes two columns in the data frame in terms
of a linear relationship. For instance `y~x` represents a model in which
$y = a + b x$. In fact, the data rarely lie exactly along a line, but 
the `lm` function will find the values for `a` and `b` that define the line
that best  relates _y_ to _x_ (specifically this will be the line such that
the minimizes the sum of the square of the errors between the actual _y_
and the line:
$\sum_i (y_i - (a + b x_i))^2$.

We will be using `lm` in this lab to find the rates of change of different
parameters in the Kaya identity. For that we want to examine the slopes of the
lines identified by `lm`. We can do that as follows for the slope that relates
wind speed to pressure for a hurricane.

First, let's plot the two variables to see wether there's an apparent 
relationship. 
Let's look at hurricanes from 2005:

```{r plot_pressure_wind, echo=TRUE, eval=TRUE, include=TRUE, warning=FALSE, message=FALSE}
storms %>% filter(status == "hurricane", year == 2005) %>%
  ggplot(aes(x = pressure, y = wind)) + geom_point(na.rm = T)
```

The data clearly doesn't lie on a straight line, but there is a clear
relationship that lower pressures tend to go along with higher wind
speeds and higher pressures with lower wind speeds.

First, let's review how we can fit a relationship between wind speed and 
pressure for hurricanes from 2005:
```r
lin_fit = storms %>% filter(status == "hurricane", year == 2005) %>%
           lm(wind ~ pressure, data = .)
```
But how can we get the slope out of the variable `lin_fit`?
Here we use the `broom` package, which gives us a `tidy` function
that will tidy up the results of statistical models and put them
into a nice data frame:

```{r tidy_fit_1, echo=TRUE, eval=TRUE, include=TRUE, warning=FALSE, message=FALSE}
lin_fit = storms %>% filter(status == "hurricane", year == 2005) %>%
           lm(wind ~ pressure, data = .)
tidy(lin_fit)
```

Here, `term` tells us the parameter: `(Intercept)` is the intercept 
(_a_ in the formula above), and `pressure` is the slope with respect to the
pressure (_b_ in the formula above).
`estimate` is the estimate of the parameter (intercept or slope),
`std.error` is the uncertainty about the estimate of the parameter,
`statistic` and `p.value` are used for statistical hypothesis testing and we
will not use them so you can ignore them.

If I just want the slope, I can get that as follows:
```{r tidy_fit_2, echo=TRUE, eval=TRUE, include=TRUE, warning=FALSE, message=FALSE}
lin_fit = storms %>% filter(status == "hurricane", year == 2005) %>%
           lm(wind ~ pressure, data = .)
slope = tidy(lin_fit) %>% filter(term == "pressure") %$% estimate
slope
```

Here, `filter` selects the row where term equals "pressure" and `%$% estimate`
then selects the `estimate` column.

We can write this more compactly as follows:
```{r tidy_fit_3, echo=TRUE, eval=TRUE, include=TRUE, warning=FALSE, message=FALSE}
slope = storms %>% 
           filter(status == "hurricane", year == 2005) %>%
           lm(wind ~ pressure, data = .) %>% 
           tidy() %>% 
           filter(term == "pressure") %$% estimate
slope
```

# Writing functions

If you are going to do the same calculation many times in R it is often useful
to define a function that performs that calculation and then you can simply
call the function whenever you want to do the calculation with new values.

For instance, in the "Growth Rates and Trends" section of the assignment
for the bottom-up decarbonization lab, we see that if we have a quantity
_q_ that is growing at a steady percentage rate $r_q$ each year and we know
the value in one year $y_0$ and want to predict what the value will be at
a future year $y$, we can calculate this as follows:
$$ q(y) = q(y_0) \times \exp(r_q \times (y - y_0)). $$
We can write this as an R function that we can call for different quantities and
years:
```{r growth_function, echo=TRUE, include=TRUE}
growth = function(q_0, rate, start_year, end_year) {
  q_0 * exp(rate * (end_year - start_year))
}
```
A function definition has three parts:
First we have the name of the function (`growth`), then we have the word
`function`, followed by the arguments that the user must supply to the function
`(q_0, rate, start_year, end_year)`, and then the body of the function, between
curly braces `{ ... }`.
When you call the function, it executes all of the code between the curly braces
and returns the last value calculated before the closing brace.
Thus, we can call the function like this to estimate the world population 
in 2050
```{r call_function, echo=TRUE, include=TRUE}
P_2017 = 7.53 # world population was 7.53 billion
r_P = 0.0141  # growth rate is 1.41% per yer
P_2050 = growth(q_0 = P_2017, rate = r_P, start_year = 2017, end_year = 2050)
P_2050
```
If you don't specify the name of the parameter, the function fills them out in
order, so we could also write 
```{r call_function_2, echo=TRUE, include=TRUE}
P_2017 = 7.53 # world population was 7.53 billion
r_P = 0.0141  # growth rate is 1.41% per yer
P_2050 = growth(P_2017, r_P, 2017, 2050)
P_2050
```

# `kayadata`: an R package for analyzing Kaya-identity data

You can install the `kayadata` package as follows:

```r
library(devtools)
install_github("jonathan-g/kayadata")
```
or
```r
library(pacman)
p_load_current_gh("jonathan-g/kayadata")
```

Once you have installed the `kayadata` package, you load it as follows:
```r
library(kayadata)
```

This package has several useful functions:

* `get_kaya_data(region_name)`: Get Kaya-identity data for a country or region.
    * **Argument:** `region_name` is the name of the country or region you want 
      data for (e.g., "United States", "World", "China", etc.).
      
        You can inspect a list of countries and regions by running 
        `kaya_region_list()`.
    * **Results:** The function returns a data frame with the following columns:
        * `region`: The country or region.
        * `year`: The year.
        * `P`: The population (in billions).
        * `G`: The gross domestic product (in trillions).
        * `E`: The primary energy consumption (in quads).
        * `F`: The energy-related CO~2~ emissions (in millions of metric tons 
          of CO~2~).
        * `g`: Per-capita GDP (in thousands of dollars per person).
        * `e`: The energy intensity of the economy (in quads per trillion 
          dollars of GDP).
        * `f`: The emissions intensity of the energy supply (in millions of 
          metric tons of CO~2~ per quad).
        * `ef`: The emissions intensity of the economy (in millions of metric 
          tons of CO~2~ per trillion dollars of GDP).
* `get_top_down_values(region_name)`: Get projections of the future
  values of Kaya-identity variables for a country, from the top-down 
  International Energy Outlook analysis by the U.S. Energy Information 
  Administration.
    * **Argument:** `region_name` is the name of the country or region you want 
      data for (e.g., "United States", "World", "China", etc.).
    * **Results:** The function returns a data frame with the following columns:
        * `region`: The country or region.
        * `year`: The year (five-year intervals from 2015--2050).
        * `P`: The population (in billions).
        * `G`: The gross domestic product (in trillions).
        * `E`: The primary energy consumption (in quads).
        * `F`: The energy-related CO~2~ emissions (in millions of metric tons 
          of CO~2~).
        * `g`: Per-capita GDP (in thousands of dollars per person).
        * `e`: The energy intensity of the economy (in quads per trillion 
          dollars of GDP).
        * `f`: The emissions intensity of the energy supply (in millions of 
          metric tons of CO~2~ per quad).
        * `ef`: The emissions intensity of the economy (in millions of metric 
          tons of CO~2~ per trillion dollars of GDP).
* `get_top_down_trends(region_name)`: Get the trends for Kaya-identity variables 
  for a country, from the top-down International Energy OUtlook analysis by the 
  U.S. Energy Information Administration.
    * **Argument:** `region_name` is the name of the country or region you 
      want data for (e.g., "United States", "World", "China", etc.).
    * **Results:** The function returns a data frame with the following columns:
        * `region`: The country or region
        * `P`: The rate of change of the population (multiply by 100 to get 
          the rate in percent per year).
        * `G`: The rate of change of GDP.
        * `E`: The rate of change of primary energy consumption.
        * `F`: The rate of change of CO~2~ emissions.
        * `g`: The rate of change of per-capita GDP.
        * `e`: The rate of change of the energy intensity of the economy.
        * `f`: The rate of change of the emissions intensity of the energy 
          supply.
        * `ef`: The rate of change of the emissions intensity of the economy.
* `project_top_down(region_name, year)`: Calculate projected values for 
  Kaya identity variables at a future year from the present through 2050.
    * **Arguments:** 
        * `region_name` is the name of the country or region you want 
          data for.
        * `year`: The year you want projected data for (through 2050).
    * **Results:** A data frame with one row and columns for `region`,
      `year`, and the Kaya variables: _P_, _G_, _E_, _F_, _g_, _e_, _f_, and
      _ef_.
* `get_fuel_mix(region_name)`: Get the mixture of primary energy sources used
  by a country or region.
      * **Argument:** `region_name` is the name of the country or region you 
      want data for (e.g., "United States", "World", "China", etc.).
    * **Results:** The function returns a data frame with the following columns:
        * `region`: The country or region
        * `region_code`: A three-letter code for the country or region.
        * `geography`: The kind of region ("nation", "region", or "world").
        * `year`: The year the data represents.
        * `fuel`: The kind of energy source ("Oil", "Natural Gas", "Coal", 
          "Nuclear", "Hydro", or "Renewables")
        * `quads`: The number of quads of that kind of energy consumed in that 
          year.
        * `pct`: The percent of total primary energy consumption that came from
          that kind of energy source.
* `plot_kaya(kaya_data, variable, start_year, stop_year, y_lab, log_scale,
   trend_line, points, font_size)`: Plot a time series showing trends in a 
   Kaya variable.
    * **Arguments:**
        * `kaya_data`: A data frame, like those returned by `get_kaya_data`.
        * `variable`: A quoted Kaya variable name ("P", "G", "E", "F", "g", 
          "e", "f", or "ef").
        * `start_year` (optional): If you want to do trend analysis, you can
          specify the year to start calculating the trend.
          The graph will color the years from `start_year` to `stop_year`
          in a darker color.
          
          If you don't specify `start_year`, the default value is 1980.
          You can eliminate the special coloring and trend analysis 
          by setting `start_year = NULL`.
        * `stop_year` (optional): The year to stop the trend analysis. 
          If you don't specify `stop_year`, it defaults to the last year
          in the data frame.
        * `y_lab` (optional): The label for the y-axis. If you don't specify 
          this, R will pick an appropriate label based on which variable you
          choose.
        * `log_scale` (optional): Set this to `TRUE` or `FALSE` to specify 
          whether to plot the y-axis on a logarithmic scale. 
          The default is to set `log_scale = FALSE` if you don't specify a 
          value.
        * `trend_line` (optional): Set this to `TRUE` or `FALSE` to specify 
          whether to plot a trend line. 
          The default is to set `trend_line = FALSE` if you don't specify a 
          value.
        * `points` (optional): Set this to `TRUE` or `FALSE` to specify 
          whether to plot points on the line. If you are making a small
          graph, the points may make the graph too busy.
          The default is to set `points = TRUE` if you don't specify a 
          value.
        * `font_size` (optional): This lets you adjust the size of the fonts
          used in labeling the axes on the graph.
          If you are making small plots and the y-axis label won't fit on
          the side of the graph, you might want to use a smaller font size.
          The default is to set `font_size = 20` if you don't specify a value.
    * **Result:** This function returns a plot object.
    * **Example:**
      ```{r example_kaya_plot, echo=TRUE, include=TRUE}
      kd <- get_kaya_data("World")
      plot_kaya(kd, "F")
      ```
      ```{r example_kaya_plot_2, echo=TRUE, include=TRUE}
      plot_kaya(kd, "F", log_scale = TRUE, font_size = 12, trend_line = TRUE)
      ```
      ```{r example_kaya_plot_3, echo=TRUE, include=TRUE}
      plot_kaya(kd, "F", points = FALSE, start_year = NULL, font_size = 12)
      ```
* `plot_fuel_mix(fuel_mix, collapse_renewables)`: Make a "donut" chart showing
  the amount of energy the country got from different energy sources.
    * **Arguments:**
        * `fuel_mix`: A data frame with the fuel mix for the country, like the
          ones returned by `get_fuel_mix()`
        * `collapse_renewables` (optional): Set this to `TRUE` or `FALSE` to
          specify whether to combine "Hydro" with other renewables to make one
          "Renewables" category (`TRUE`), or plot "Hydro" separately from the
          other renewables (`FALSE`).
          
          If you don't specify `collapse_renewables`, the default is to set
          `collapse_renewables = TRUE`.
    * **Results:** A plot object with a donut chart.
    * **Example:**
      ```{r example_fuel_plot, echo=TRUE, include=TRUE}
      fm <- get_fuel_mix("World")
      plot_fuel_mix(fm)
      ```
      ```{r example_fuel_plot_2, echo=TRUE, include=TRUE}
      plot_fuel_mix(fm, collapse_renewables = FALSE)
      ```
* `kaya_region_list()`: Get a list of regions and countries for which you can
  get Kaya data.
* `emissions_factors()`: Get a data frame with columns for 
    * `fuel`: The energy source
    * `emission_factor`: The emissions from that energy source, in million
      metric tons of CO~2~ per quad of energy.
* `generation_capacity()`: Get a data frame with generation capacity of a 
  typical power plant for different kinds of energy sources. 
  This data frame has columns:
    * `fuel`: The energy source.
    * `description`: A description of the kind of power plant.
    * `nameplate_capacity`: The "nameplate capacity" of a typical power plant
      of that type. The nameplate capacity is the power output when the
      plant is running at its maximum capacity.
    * `capacity_factor`: The capacity factor. This is the fraction of the 
      nameplate capacity that the power plant can produce over an average 
      year.
      
        The average power output from a power plant is the nameplate capacity
        times the capacity factor:
        $$ \text{average power} = \text{nameplate capacity} \times 
           \text{capacity factor} $$

